---
title: "Data exploration assignment"
author: "Serena Wu"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

In this exercise, you will practice basic `dplyr` (the relevant package
within `tidyverse`) operations on a dataset of gene expression levels
across human tissues (from the GTEx Project). The dataset (`dt`) has one
row per gene, with columns:

- `Name`: the Ensembl gene ID
- `Description`: the gene symbol
- One column per tissue containing the median expression levels (in TPM
  units)

**Using HopGPT.** You are welcome to use HopGPT on this assignment. If
you do, add a couple of comment lines in that same code block noting
what you asked it and how you checked the answer. Rmember that a chatbot
will happily give you code that runs perfectly and answers a slightly
different question than the one you asked.

For example:

```{r ai-comment-example, eval=FALSE}
# Asked HopGPT how to sort a tibble by a column with a space in its name.
# Checked: printed the first and last rows to confirm it sorted descending.
```

------------------------------------------------------------------------

### Start by loading the data into R:

Note the arguments, which specify that the first two lines of the file
will be skipped, that the file contains a header line that should be
used to form column names, and that the columns are tab (`\t`)
separated.

```{r load-data}
library(tidyverse)

dt <- read_delim("~/data/GTEX/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz",
                 skip = 2, col_names = TRUE, delim = "\t")
```

### Q1. How many genes are in the dataset, and how many tissues are represented?

Note that not every column is a tissue.

```{r q1}

# code and answer here
# how many genes 
length(dt$Name)
ncol(dt)-2
```

------------------------------------------------------------------------

### Q2. How many genes are expressed above a given threshold in each tissue?

A common threshold is 1 TPM to define whether a gene is "expressed."\
Compute, for each of the three tissues, `Liver`, `Testis`, and
`Whole Blood`, how many genes are expressed above this threshold.\
Then identify which tissue has the **most** and which has the **fewest**
expressed genes. See this paper for one possible explanation: Xia et al.
2021: [PMC7891839](https://pmc.ncbi.nlm.nih.gov/articles/PMC7891839/).

Do the results change if you set the threshold higher, say at 10 TPM?

*Hint:* the column name `Whole Blood` contains a space, so it needs
backticks: `` dt$`Whole Blood` ``.

```{r q2}

# Code here
nrow(filter(dt,Liver >1))
nrow(filter(dt,Testis >1))
nrow(filter(dt,'Whole Blood'>1))
nrow(filter(dt,Liver >10))
nrow(filter(dt,Testis >10))
nrow(filter(dt,'Whole Blood'>10))

# Interpretation here as a comment
#Liver has 13499 expressed genes, while Testis has 25064 and Whole Blood has 56200. That tells us that Whole Blood has the most expressed genes and Liver has the fewest expressed genes. This trend is the same if we changet he threshold higher to 10 TPM that whole blood still has the most expressed genes while liver has the least. 
```

------------------------------------------------------------------------

### Q2b. Write a function

If you answered Q2 by copying the same line of code and editing the
tissue name, and then copying those lines again and editing the
threshold, you did six near-identical things by hand. Each time you
copied code represents a chance to introduce a mistake, and if you
wanted to change something, you would have to implement the change
multiple times.

Write a function `count_expressed()` that takes a vector of expression
values and a threshold, and **returns** the number of genes above that
threshold. Give the threshold a default value of 1.

Test it on a small vector where you already know the answer, such as
`c(0, 0.5, 1, 5, 100)`. Then use it to redo Q2 (all three tissues, both
thresholds) instead of writing out the counting code separately six
times.

```{r q2b}

# your function here
count_expressed <- function(name, threshold){
  count <- nrow(filter(dt,name >threshold))
  return(count)
}

# test on an input with a known answer
count_expressed(dt$Liver, 1)

# use it to answer Q2 again
#count_expressed(dt$Liver)
count_expressed(dt$Testis, 1)
count_expressed(dt$`Whole Blood`,1)

```

------------------------------------------------------------------------

### Q3. What are the 20 most highly expressed genes in the liver?

What do you notice about many of the genes in this list? Can you propose
a hypothesis to explain your observation?

```{r q3}

# Code here
arrange(dt,desc(Liver)) %>% 
  head(20) %>% select(c(Description, Liver))
# Interpretation here as a comment
#I notice some of the highly expressed genes in the liver are from "MT" or mitochrondria. A hypothesis could be if we can get a highly expressed gene from the liver from an experiment, there is a high chance that the gene is from mitochrondria. 
```

------------------------------------------------------------------------

### Q4. Compare the mean expression levels of mitochondrial versus nuclear-encoded genes in the Liver.

Genes encoded by the mitochondrial genome have the gene symbol prefix
`MT-`. The function `str_starts()` from the `stringr` package in
`tidyverse` tests whether a string (or a vector of strings) starts with
a given pattern, returning TRUE/FALSE (or a boolean vector) as output.

Use `str_starts()` to identify which genes are mitochondrially encoded
and store this information in a new Boolean column named
`is_mitochondrial`. Next, compute the mean and median expression of
mitochondrial genes in the liver, as well as the mean and median
expression of nuclear-encoded genes in the liver and compare them. Do
mitochondrial genes exhibit higher or lower mean and median expression
than nuclear genes? What does the difference between the mean and median
tell you about the distribution of gene expression?

**Check your pattern before you trust the result.** There is an easy
version of this that is wrong: matching `"MT"` instead of `"MT-"`. Count
how many genes each of the two patterns matches, and look at a few of
the genes that `"MT"` picks up but `"MT-"` does not. How much does the
mistake change your answer?

```{r q4}
# Code here
#how many genes start with "MT-" 
dt <- dt %>% 
  mutate(is_mitochondrial = str_starts(Description, "MT-")) 
dt_mito <- filter(dt,is_mitochondrial== T) 
nrow_dt_mito <- nrow(dt_mito)
nrow_dt_nonmito <- nrow(dt_nonmito)
dt_nonmito <- filter(dt,is_mitochondrial== F) 
#mean 
print(mean(dt_mito$Liver))
print(median(dt_mito$Liver))
print(mean(dt_nonmito$Liver))
print(median(dt_nonmito$Liver))

print(nrow_dt_mito)
# how many genes match "MT-" vs "MT"? which genes differ?
# 37 genes are mitochrondrial and 56163 are not mitochrondrial. 
             
# Interpretation here as a comment
#mean for mitochrondrial genes are a lot higher than the median, and the reason could be the outiers are changing or increasing the mean value a lot. the mean for non-mitochrondrial genes is much lower and it is similar to the median of the non-mitochrondrial genes which is expected. 
```

------------------------------------------------------------------------

### Q5. Which tissue is this?

So far we have told you which tissue to look at. Now you will find a
tissue based on the genes it expresses.

A gene is *tissue-specific* if it is expressed much more in one tissue
than it is on average across the body. To measure that, you first need
each gene's average expression across all tissues.

```{r mean-across-tissues}
# remove the columns that are not tissue expression
tissue_only <- dt %>%
  select(-Name, -Description, -is_mitochondrial)

# store the mean back in the original tibble
dt <- dt %>%
  mutate(mean_all_tissues = rowMeans(tissue_only))

dt %>%
  mutate(specificity = log2((Liver + 1) / (mean_all_tissues + 1))) %>%
  filter(Liver > 10) %>%
  arrange(desc(specificity)) %>%
  select(Description, Liver, mean_all_tissues, specificity) %>%
  head(20)
```

**(a)** Pick any tissue in the dataset and compute a specificity score
for every gene:

$$\text{specificity} = \log_2 \left( \frac{\text{my tissue} + 1}{\text{mean across all tissues} + 1} \right)$$

Keep only genes expressed above 10 TPM in your tissue, sort by
specificity, and print the top 20 gene symbols.

```{r answer 5b}
# remove the columns that are not tissue expression
tissue_only <- dt %>%
  select(-Name, -Description, -is_mitochondrial)

# store the mean back in the original tibble
dt <- dt %>%
  mutate(mean_all_tissues = rowMeans(tissue_only))

dt %>%
  mutate(specificity = log2((Liver + 1) / (mean_all_tissues + 1))) %>%
  filter(Liver > 10) %>%
  arrange(desc(specificity)) %>%
  select(Description) %>%
  head(20)

#comment: if we only look at the gene symbol that are highly expressed, we may not be able to get too much info about the tissue from just the gene symbol because we don't know which gene is from which tissue. 
```

**(b)** Give a classmate *only that list of 20 gene symbols* — not the
tissue name — and see whether they can identify the tissue. Do the same
for theirs. If the list doesn't give it away, explain why it
challenging.

**(c)** Why is the `> 10 TPM` filter in part (a)? Take it out and look
at what rises to the top. What kind of genes are the most \`\`specific"
when you do not require them to be meaningfully expressed?

```{r answer 5c}
# remove the columns that are not tissue expression
tissue_only <- dt %>%
  select(-Name, -Description, -is_mitochondrial)

# store the mean back in the original tibble
dt <- dt %>%
  mutate(mean_all_tissues = rowMeans(tissue_only))

dt %>%
  mutate(specificity = log2((Liver + 1) / (mean_all_tissues + 1))) %>%
  arrange(desc(specificity)) %>%
  select(Description, Liver, mean_all_tissues, specificity) %>%
  head(20)

#comment: for the code in part a, I removed the threshold 10 TPM and the output or the top 20 genes look similar to part a. this may be because the top genes are already highly liver specific. 
```

**(d)** Housekeeping genes, which are highly expressed in every tissue,
should score near 0 on this specificity metric. Find a few genes with
specificity closest to 0 and high expression. Do they look like what you
would expect?

```{r answer 5d}

# remove the columns that are not tissue expression
tissue_only <- dt %>%
  select(-Name, -Description, -is_mitochondrial)

# store the mean back in the original tibble
dt <- dt %>%
  mutate(mean_all_tissues = rowMeans(tissue_only))

# Asked GPT how to calculate the specificity score for housekeeping genes
dt %>%
  mutate(specificity = log2((Liver + 1) / (mean_all_tissues + 1))) %>%
  filter(Liver > 10) %>%
  arrange(abs(specificity)) %>%
  select(Description, Liver, mean_all_tissues, specificity) %>%
  head(20)

#comment: the housekeeping genes like MT-NDS are highly expressed in liver and all tissues, but the specificity is very close to 0. this is what I expect for housekeeping genes that the specificity score is low for all specific tissue. 

```

------------------------------------------------------------------------

### Q6. Comparing gene expression between tissues

Another way to compare tissues is to ask how similar are their
expression patterns overall, across all genes?

R has a built-in function called `cor.test()` that measures the
correlation between two vectors of numbers. Remember that you can
extract the contents of a given column of a tibble as a vector using the
syntax `tibble_name$column_name`. If a column name contains spaces,
enclose the column name in backticks such as
`` tibble_name$`column name` ``.

Use `cor.test()` with the argument `method = "spearman"` to compute:

- the Spearman correlation between `Liver` and `Heart - Left Ventricle`
- the Spearman correlation between `Liver` and `Brain - Cortex`

Which pair of tissues is more strongly correlated? Does this result make
sense biologically?

You will see a warning about not being able to compute an exact p-value
because of ties. You can ignore it, as it does not affect the
qualitative results in this case. It is worth understanding where it
comes from, though: what is it about these data that creates so many
tied values?

```{r q6}
# Liver vs Heart answer
cor.test(dt$Liver,dt$`Heart - Left Ventricle`, method = "spearman")
# Liver vs Brain answer
cor.test(dt$Liver,dt$`Brain - Cortex`, method = "spearman")
# interpretation
#comparing these two pairs of tissues, it seems like heart is slightly more correlated than the brain and it makes sense bioloically because as we already know, brain is a very special organ and it can do things in a different way.
```

------------------------------------------------------------------------

### Optional: further exploration

What other patterns are present in the data? Use tidyverse functions to
explore and summarize the data in different ways and see what you can
learn.

```{r optional}
# answer here
```
