---
title: "Data visualization assignment"
author: "Serena Wu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE, fig.width = 6, fig.height = 4)
```

In this exercise, you will practice basic plotting with **`ggplot2`** on a
dataset of median gene expression levels across samples from various human
tissues from the GTEx Project. The dataset (`dt`) has one row per gene, with
columns:

-   `Name`: the Ensembl gene ID
-   `Description`: the gene symbol
-   One column per tissue containing the median expression levels (in TPM units)

Insert your code into the empty code blocks. Insert answers to any comprehension
questions as comments within the code blocks. You do not need to submit the
figures themselves, only the code used to produce them. Remember to label your
axes!

**Using HopGPT.** As in the last assignment: you are welcome to use it, and
wherever you do, add a couple of comment lines in that same code block noting
what you asked it and how you checked the answer.

------------------------------------------------------------------------

### Load data

Start by loading the data into R, as before.

```{r load-data}
library(tidyverse)

dt <- read_delim("~/Data/GTEX/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz",
                 skip = 2, col_names = TRUE, delim = "\t")
```

Recreate the `is_mitochondrial` column from the last assignment; you will need it
in Q4 and Q5.

```{r add-mito}
dt <- dt %>%
  mutate(is_mitochondrial = str_starts(Description, "MT-"))

dt %>% count(is_mitochondrial)
```

------------------------------------------------------------------------

### Q1. Histogram

Make a histogram of gene expression in the liver. Use the argument
`breaks = seq(0, 100, 2)` to bin in units of 2 TPM, ranging from 0 to 100. Note
that this cuts off the data at 100, which might be misleading depending on what
we are trying to depict! How many genes are you leaving out?

What do you notice about the shape of the distribution? What does it tell you
about the relative expression of genes in the liver transcriptome? Do you think
this is specific to the liver? Check another tissue, as well.

```{r q1}
ggplot(dt,aes(x= Liver))+
  geom_histogram(breaks = seq(0, 100, 2))+
  labs(x = "Expression(TPM)", y = "Number of Genes")

#check historgram pattern with bladder 
ggplot(dt,aes(x= Bladder))+
  geom_histogram(breaks = seq(0, 100, 2))+
  labs(x = "Expression(TPM)", y = "Number of Genes")
#comment: right now, we are leaving out the genes that have TPM that is great than 100. we are missing out 594 genes based on the code below. I notice the shape of the distribution is right skewed. It tells me that the relative expression of genes in the liver is higher at low TPM I check the distribution of another tissue bladder and the output looks similar to the liver distribution. So this pattern is not specific to liver. 
# number of genes > 100 TPM 
dt |>
  filter(Liver > 100) |> 
  nrow()
# answer here
```

------------------------------------------------------------------------

### Q2. Density plot

Going back to the liver, now represent the same data with a density plot rather
than a histogram. Do not restrict the X axis. Within the same code block, create
a second plot where you remove the custom bin breaks, but transform the expression
values using `log10()`. First add a pseudo-count of 1 to the expression values, as 
the log of 0 is negative infinity.

What does the log-transformed version show you that the first plot does not? And where
on the log-transformed plot do the genes with 0 TPM end up?

```{r q2}
# answer here
dt %>% 
  ggplot(aes(x= Liver))+
  geom_density()+
  labs(x = "Expression(TPM)", y = "Density")
#second plot with 
dt %>% 
  ggplot(aes(x= log10(Liver+1)))+
  geom_density()+
  labs(x = "Expression(log10(TPM+1))", y = "Density")



```

------------------------------------------------------------------------

### Q3. Scatter plot

Make a scatter plot comparing the expression of genes in the `Liver` to the same
genes in the `Heart - Left Ventricle`. Add a pseudocount to each and log10 transform
them. Adjust the point size to 1 and the alpha (opacity) to 0.2. Add a dashed red line 
along the diagonal (y = x).

What does a point far above the diagonal mean, biologically?

```{r q3}

# answer here
dt %>% 
  ggplot(aes(x= log10(Liver+1), y = log10(`Heart - Left Ventricle`+1))) +
  geom_point(size = 1, alpha = 0.2) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed', color = 'red')+
  labs(x = "Liver Expression (log10(TPM+1))", y = 'Heart-Left Ventricle Expression(log10(TPM+1))')
#comment: the point above the diagonal means it is higher on the y axis, which mean higher expression in the heart than in the liver. on the other hand, the points below the diagonal line are more highly expressed in liver than in the herat. 
```

------------------------------------------------------------------------

### Q4. Colored points on scatter plot

Color the previous scatter plot points based on whether the genes are
mitochondrial vs. nuclear encoded, using the `is_mitochondrial` column you made
above.

There are only a few dozen mitochondrial genes among tens of thousands of
nuclear ones, so the colored points end up buried. Try plotting them as a second
`geom_point()` layer on top of the others, using the `data` argument to give that
layer just the mitochondrial genes:
`geom_point(data = filter(dt, is_mitochondrial), ...)`.

Where do the mitochondrial genes sit relative to the diagonal, and does that fit
what you found in the last assignment?

```{r q4}

# answer here
dt %>% 
  ggplot(aes(x= log10(Liver+1), y = log10(`Heart - Left Ventricle`+1))) +
  geom_point(size = 1, alpha = 0.2, color = "black") +
  geom_point(data = filter(dt, is_mitochondrial), size = 1, alpha = 0.2, color = "green") +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed', color = 'red')+
  labs(x = "Liver Expression (log10(TPM+1))", y = 'Heart-Left Ventricle Expression(log10(TPM+1))')
#comment: the few mitochondrial genes are mostly close to the diagonal line which suggests their liver and heart expression are relatively similar. Most of them are slighly higher than the diagonal line which suggest a slighly higher expression in heart since y axis is the heart expression.

```

------------------------------------------------------------------------

### Q5. Facets

Repeat question 1 (histogram), but faceting such that you create separate panels
(arranged horizontally) for several tissues of interest:

```{r focal-tissues}
focal_tissues <- c("Liver", "Testis", "Whole Blood", "Heart - Left Ventricle", "Brain - Cortex")
```

This requires first converting the data into "long" format with `pivot_longer()`:

```{r pivot-long}
dt_long <- dt %>%
  select("Description", "is_mitochondrial", all_of(focal_tissues)) %>%
  pivot_longer(cols = all_of(focal_tissues), names_to = "Tissue", values_to = "TPM")
```

After creating the 5-panel plot, also split into two rows that separate genes
that are versus are not encoded by the mitochondrial genome (10 panels). Use the
argument `scales = "free_y"` to deal with the fact that there are far fewer
mitochondrial-encoded than nuclear-encoded genes.

`scales = "free_y"` means the panels can no longer be compared by eye. Write the
one-sentence note you would put in a figure caption to keep a reader from being
misled.

```{r q5}

# your code for 5-panel plot
dt_long %>% 
  ggplot(aes(x= TPM, fill = Tissue)) +
  geom_histogram(breaks = seq(0,100,2)) +
  facet_wrap(~Tissue, nrow = 1) +
  labs(x= 'Expression (TPM)',y = 'Number of genes') +
  theme(legend.position = "none")
  
# your code for 10-panel plot
dt_long <- dt_long %>% 
  mutate(gene_origin = if_else(is_mitochondrial, "Mitochondrial-encoded", "Nuclear-encoded"))
dt_long %>% 
  ggplot(aes(x= TPM, fill = Tissue)) +
  geom_histogram(breaks = seq(0,100,2)) +
  facet_grid(gene_origin ~Tissue, scales = "free_y") +
  labs(x= 'Expression (TPM)', y = 'Number of genes') +
  theme(legend.position = "none")


# caption note
#use GPT to help me understand scales = "free_y" and how it helps with the visualization in this case 
#using scales = "free_y" allows each panel to choose its own y axis because the number of genes between mitochondrial and non-mitochondrial are very different. so the heights of the bars between mitochondrial and non-mitochondrial genes are not meant to be compared directly. 
```

------------------------------------------------------------------------

Optional: Have fun and explore these data! Create your own plots and interpret
them.

```{r optional}

# answer here
```
