# Linear Regression

**Deadline:** Friday, October 9\
**Resubmission deadline:** Friday, October 30

## Assignment Overview

The goal of today's lab is to use linear regression and related statistical methods to investigate the relationship between paternal age, maternal age, and the number of de novo mutations (DNMs) in a proband (offspring). Today's assignment will build familiarity with manipulating tabular datasets containing mixed data types using the **tidyverse** in **R**. Specifically, you will import a table of de novo mutations and manipulate it to calculate the number of maternal and paternal DNMs per individual. You will then fit and interpret simple and multiple linear regression models with **stats::lm**, compare nested models with **stats::anova**, and tidy results with **broom**.

This assignment is an R Markdown notebook. Write your code in the empty code chunks and your answers where you see *Your answer:*.

## Data

Data are taken from [Halldorsson, B. V. et al. (2019). Characterizing mutagenic effects of recombination through a sequence-level genetic map. *Science*, 363(6425)](https://science.sciencemag.org/content/363/6425/eaau1043.abstract).

Read the abstract from the above paper to understand the context of the datasets you will be using. The data you need for this assignment are available from Dropbox at:

1.  Information about the number and parental origin of each de novo mutation detected in a proband (offspring): [aau1043_dnm.csv](https://www.dropbox.com/scl/fi/6e28a3dow872fi02cp537/aau1043_dnm.csv?rlkey=l3gs7fb6igff4el5ov6wg96ai&dl=0)
2.  Ages of the parents of each proband: [aau1043_parental_age.csv](https://www.dropbox.com/scl/fi/sjrq1x1g30h0j10ktxysi/aau1043_parental_age.csv?rlkey=e9g3m9iq4tfsm9ski7w4vb0bf&dl=0)

You may copy these into your submission directory (and add to your `.gitignore`).

Before beginning the assignment, take a quick look at both files (e.g., with `less -S` in Unix) to confirm their structure.

------------------------------------------------------------------------

## Getting started (R packages)

Load the tidyverse and broom packages.

```{r}
library(tidyverse)
library(broom)
```

------------------------------------------------------------------------

## Exercises

### Exercise 1: Wrangle the data

#### **Step 1.1 — Load DNMs**

Load `aau1043_dnm.csv` into a tibble.

```{r}
data <- read_csv("~/Downloads/aau1043_dnm.csv")
head(data)
```

#### **Step 1.2 — Count DNMs by parental origin per proband**

Create a **per-proband** summary with counts of maternally and paternally inherited DNMs. Ignore DNMs without a specified parent of origin. Name the count columns `maternal_dnm` and `paternal_dnm`, as later steps use these names.

```{r}
data_count <- data |>
  filter(Phase_combined == "mother" | Phase_combined == "father") |>
  group_by(Proband_id) |>
  summarise(
    maternal_dnm = sum(Phase_combined == "mother"),
    paternal_dnm = sum(Phase_combined == "father")
  )

```

#### **Step 1.3 — Load parental ages**

Load `aau1043_parental_age.csv`.

```{r}
age <- read_csv("~/Downloads/aau1043_parental_age.csv")
```

#### **Step 1.4 — Merge counts with ages**

Join the two tibbles by proband ID. Name the result `merged`, as later steps use this name.

```{r}
merged <- left_join(data_count, age, by = "Proband_id")
```

------------------------------------------------------------------------

### Exercise 2: Fit and interpret linear regression models with R

Use your merged data frame for the following. All plots should be clearly labeled and easily interpretable.

#### **Step 2.1 — Visualize relationships**

**2.1.1** Create a scatter plot of the count of maternal DNMs vs. maternal age → save as `ex2_a.png`

```{r}
colnames(merged)
A <- ggplot(merged, aes(x = Mother_age, y = maternal_dnm)) +
  geom_point() +
  labs(x = "Maternal age",
       y = "Maternal DNMs")

ggsave("ex2_a.png", A)
```

**2.1.2** Create a scatter plot of the count of paternal DNMs vs. paternal age → save as `ex2_b.png`

```{r}
B <- ggplot(merged, aes(x = Father_age, y = paternal_dnm)) +
  geom_point() +
  labs(x = "Paternal age",
       y = "Paternal DNMs")

ggsave("ex2_b.png", B)
```

**2.1.3** Create a scatter plot of paternal age vs. maternal age → save as `ex2_c.png`

```{r}
C <- ggplot(merged, aes(x = Mother_age, y = Father_age)) +
  geom_point() +
  labs(x = "Maternal age",
       y = "Paternal age")

ggsave("ex2_c.png", C)

getwd()
```

#### **Step 2.2 — OLS: maternal age vs. maternal DNMs**

Fit a simple linear regression model relating maternal age to the number of maternal de novo mutations.

```{r}
maternal_model <- lm(maternal_dnm ~ Mother_age, data = merged)

summary(maternal_model)
```

Answer the following questions:

1.  What is the "size" (i.e., slope) of this relationship? Interpret the slope in plain language. Does it match your plot?

    *Your answer: The slope is 0.37757 which means the chance of mutation increases as mothers age. This make sense as we age and this match the graph that it has a linear increase for DNMs as maternal age increases.*

2.  Is the relationship significant? How do you know? Explain the p-value in plain but precise language.

    *Your answer: Yes, the relationship is significant because the p value is really low and it is less than 0.05. the p value means if maternal age has no relationship with DNM, the change of observing this relationship due to random variation is 2.2e-16 which is really low.*

#### **Step 2.3 — OLS: paternal age vs. paternal DNMs**

Repeat the step above but for paternal age vs. paternal DNMs.

```{r}
paternal_model <- lm(paternal_dnm ~ Father_age, data = merged)

summary(paternal_model)
```

Answer the following questions:

1.  What is the "size" (i.e., slope) of this relationship? Interpret the slope in plain language. Does it match your plot?

    *Your answer: the slope is 1.35384 which is higher than maternal relationship slope. the slope means the number of paternal DNMs increase by 1.35 mutations per year. it matches with the scatter plot which also has a linear increase plot.*

2.  Is the relationship significant? How do you know? Explain the p-value in plain but precise language.

    *Your answer: the relationship is significant because the p value is also very low and less than 0.05. the p value means if paternal age has no relationship with DNM, the change of observing this relationship due to random variation is 2e-16 which is really low.*

#### **Step 2.4 — Predict for a 50.5-year-old father**

Use the paternal regression model to predict the expected number of paternal DNMs for a father of age 50.5. You are welcome to do this manually or using a built-in function, but show your work in the code chunk.

```{r}
predict(paternal_model, newdata = data.frame(Father_age = 50.5))
```

*Your answer:* the expected number of paternal DNMs is 78.7 for a father of age 50.5

#### **Step 2.5 — Multiple regression: correlated predictors**

Maternal DNMs arose in the mother's germline, so there is no obvious reason that the father's age should affect them.

**2.5.1** Fit a simple linear regression model relating **paternal** age to the number of **maternal** DNMs.

```{r}
paternal_maternal_model <- lm(maternal_dnm ~ Father_age, data = merged)

summary(paternal_maternal_model)
```

1.  Is paternal age a significant predictor of maternal DNMs on its own? Using your plot from 2.1.3, explain why this might happen.

    *Your answer: yes,* paternal age is a significant predictor of maternal DNMs because the p value is less than 0.05. This could happen because from 2.1.3, maternal age and paternal age are correlated and maternal age is also correlated to maternal DNMs.

**2.5.2** Fit a multiple linear regression model with both maternal age and paternal age as predictors of maternal DNMs, just as we added year as a covariate in the penguin example.

```{r}
both_model <- lm(maternal_dnm ~ Mother_age + Father_age, data = merged)

summary(both_model)
```

2.  What happened to the paternal age coefficient and its p-value?

    *Your answer: the coefficient changed from 0.23266 to -0.06665 and the pvalue changed from* 1.37e-13 to 0.18773. this suggest that paternal age is no longer a significant predictor of maternal DNMs

3.  Interpret the maternal age coefficient in plain language. How does its meaning differ from the slope you estimated in Step 2.2? (Hint: what is being held constant?)

    *Your answer: the coefficient changed from 0.37757 from step 2.2 to 0.44498 and it means that it has a stronger association when while we control for paternal age.*

**2.5.3** Use `anova()` to compare your two-predictor model from 2.5.2 to your model from Step 2.2 (maternal age only).

> **Hint:** In the live coding, we used `anova()` to test whether island improved a model that already contained sex and year. The same approach works for any pair of models where the smaller model is the larger model with some terms removed (these are called "nested" models). Here, the smaller model is the one that leaves out paternal age.

```{r}
anova(maternal_model, both_model)
```

4.  Does adding paternal age improve the model? Compare the p-value from `anova()` to the p-value for paternal age in the `summary()` of the two-predictor model. Why are they the same here, when for island we needed `anova()` to get a single answer? (Hint: how many coefficients did island add to the model, and how many does paternal age add?)

    *Your answer:* adding paternal age does not improve the model because p value is 0.1877 the same and greater than 0.05. They are the same because paternal age adds only one cofficient to the model. adding island helped because that example has multiple coefficient and they were useful for getting different slopes on different islands.

#### **Step 2.6 — Compare distributions of maternal vs. paternal DNMs**

Plot both distributions on the **same axes** as semi-transparent histograms; save as `ex2_d.png`.

```{r}
D <- ggplot(merged) +
  geom_histogram(aes(x = maternal_dnm, fill = "Maternal"),
                 alpha = 0.5, position = "identity") +
  geom_histogram(aes(x = paternal_dnm, fill = "Paternal"),
                 alpha = 0.5, position = "identity") +
  labs(x = "DNM count",
       y = "Frequency",
       fill = "Parent")

ggsave("ex2_d.png", D)
```

#### **Step 2.7 — Statistical test: maternal vs. paternal DNMs per proband**

We have **paired** observations per proband (maternal vs. paternal). The paired t-test assumes that the within-pair differences are approximately normally distributed.

**2.7.1** Apply a paired t-test in R using `t.test(merged$maternal_dnm, merged$paternal_dnm, paired = TRUE)`.

```{r}
t.test(merged$maternal_dnm,
       merged$paternal_dnm,
       paired = TRUE)
```

1.  What is the "size" of this relationship (i.e., the average difference in counts of maternal and paternal DNMs)? Interpret the difference in plain language. Does it match your plot?

    *Your answer: the mean difference is -39.23485 which means that each proband has about 39 more paternal DNMS than maternal DNMs. it matches the plot because the paternal bars are much higher than maternal bars with very overlapping at 30 DMNs.*

2.  Is the relationship significant? How do you know? Explain the p-value in plain but precise language.

    *Your answer: yes the relationship is significant because the p value is 2.2e-16 and is less than 0.05. the p value means if paternal DNM has no relationship with maternal DNM, the change of observing this relationship due to random variation is 2.2e-16 which is really low.*

**2.7.2** The paired t-test is equivalent to using the difference between the maternal and paternal DNM counts per proband as the response variable and fitting a model with only an intercept term (indicated with `1` on the right side of the model formula). Fit this model using `lm()`.

```{r}
merged <- merged |>
  mutate(data_difference = maternal_dnm - paternal_dnm)
difference_model <- lm(data_difference ~ 1, data = merged)

summary(difference_model)
```

3.  How do the results compare to the paired t-test? How would you interpret the coefficient estimate for the intercept term?

    *Your answer: the slope is the same compared to the paired t-test and both have very low p value. the intercept cofficient means that about 39 more paternal DNMs than the maternal DNMs.*

------------------------------------------------------------------------

### Exercise 3: Explore a new dataset

#### **Step 3.1 — Pick a TidyTuesday dataset**

Choose a dataset from the bottom of the [TidyTuesday README](https://github.com/rfordatascience/tidytuesday) and load it.

```{r}
data2 <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-02-10/schedule.csv')
colnames(olympics)
head(data2)

```

1.  Which dataset did you choose?

    *Your answer:* [The 2026 Winter Olympics!](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-02-10/readme.md)

#### **Step 3.2 — Explore and visualize**

Generate figures; save them as `ex3_<something>.png`.

```{r}
data2 <- data2 |>
  mutate(
    duration = as.numeric(
      difftime(end_datetime_local, start_datetime_local, units = "mins")
    )
  )

A <- ggplot(data2, aes(x = is_training, y = duration, fill = is_medal_event)) +
  geom_boxplot() +
  labs(
    x = "Training event",
    y = "Duration (minutes)",
    fill = "Medal event"
  )

A

ggsave("ex3_duration.png", A)
```

1.  What interesting patterns do you notice?

    *Your answer: an interesting pattern i see is medal events have a lower duration time generally that means usually metal events finished quicker than non-medal preliminary events. but the outliers could also be the games that require longer duration.*

#### **Step 3.3 — Pose and test a linear-model hypothesis**

1.  State a hypothesis that you can test with a linear model.

    *Your answer:* i hypothesize that event duration is associated with whether an event is a medal event and whether it is a training event.

Fit a linear model with at least two predictors to test your hypothesis, and evaluate its fit.

```{r}
medal <- lm(duration ~ is_medal_event + is_training, data = data2)

summary(medal)
```

2.  Why did you include each predictor?

    *Your answer: i included is_medal_event bc i see the trend that medal events tend to finish quicker than non-metal event. i included is_training because i think training usually take longer or could take longer when they are not in an actual competition and may have different duration time.*

3.  Report and interpret your results. How well does the model fit?

    *Your answer: both predictors are significantly associated with event duraction with p values lower than 0.05. training events is 28 minute shorter than non-training event. medal events are 38 min quicker than non-medal events non training events. but the R squred is very low only 6% of the variation in the event duration that i haven't included a more significant predictor like island that can explain the differences in duration better.*

------------------------------------------------------------------------

## Submission

- Your completed R Markdown notebook (`.Rmd`), with code in each chunk and your answers to all questions.
- The figures you saved (`ex2_a.png` through `ex2_d.png`, and your `ex3_` figures).

## Grading Rubric (Total = 10 points)

### **Exercise 1 — Wrangle the Data (2 points)**

- Load and inspect DNM data (**0.5 pt**)
- Create per-proband maternal and paternal DNM counts (**0.5 pt**)
- Load and inspect parental age data (**0.5 pt**)
- Join counts with ages into a merged table (**0.5 pt**)

### **Exercise 2 — Fit and Interpret Linear Models (6 points)**

- Step 2.1: Scatter plots for DNMs vs. parental age and paternal vs. maternal age (**1 pt**)
- Step 2.2: Fit and interpret maternal OLS model (**1 pt**)
- Step 2.3: Fit and interpret paternal OLS model (**0.5 pt**)
- Step 2.4: Predict paternal DNMs for age 50.5 (**0.5 pt**)
- Step 2.5: Multiple regression with both parental ages, and nested model comparison with `anova()` (**1.5 pts**)
- Step 2.6: Plot distributions of maternal vs. paternal DNMs (**0.5 pt**)
- Step 2.7: Paired t-test (`t.test` and `lm(diff ~ 1)`) and interpret results (**1 pt**)

### **Exercise 3 — Explore a New Dataset (2 points)**

- Step 3.1: Choose and document TidyTuesday dataset (**0.5 pt**)
- Step 3.2: Produce exploratory figure(s) (**0.5 pt**)
- Step 3.3: Pose and test a multiple linear regression hypothesis and interpret results (**1 pt**)

**Total Points: 10**

------------------------------------------------------------------------
